Two applications of logic programming to Coq

### 2020-11-07 papers/types2020/conc.tex
happy about sec 3 and 4

This paper follows on a line of research  starting in the late 80 and gaining more steam in the last five years that advocates the usefulness of proof theory and higher-order logic programming  for the many tasks around the development, enrichment and even formal verification of proof assistants. The development of the Elpi-Coq plug-in  has made this connection tighter.

>>>>

This paper follows on a line of research starting in the late 80's and
gaining more steam in the last five years, which advocates the
usefulness of proof theory and higher-order logic programming for the
many tasks concerning the development, enrichment and even formal
verification of proof assistants. The development of the Elpi-Coq
plug-in has made this connection tighter.

### 2020-11-07 papers/types2020/pbt.tex
happy about sec 3 and 4

We have presented in a previous paper~\cite{blanco19ppdp} a
proof-theoretical reconstruction of property-based testing (PBT) \cite{fink97sen}of
relational specifications, adopting techniques from foundational proof
certificates to account for several features of this testing
paradigm: from various \emph{generation} strategies, to
\emph{shrinking} and fault localization.

@@

While nothing prevents us from porting all the PBT features we have
accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
will implement only FPC corresponding to different flavors
\emph{exhaustive} generation, as adopted, e.g., in
SmallCheck~\cite{smallcheck} and
$\alpha$Check~\cite{cheney_momigliano_2017}, and their
combination. Note however that it would take  no more than two lines of
code in the \lsti{decideE} expert to implement a form of random data
generation in the sense of randomized
backtracking~\cite{pltredexconstraintlogic}.

@@

% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat DM
% LocalWords:  tif ttrue tzero tsucc SUT PBT SmallCheck QuickChick
% LocalWords:  decidability shrinkers dep pbt fpc interp forall ness
% LocalWords:  ord unfoldings pre relationally shrinker booleans typ
% LocalWords:  tfalse tpred tiszero TBool TNat subexp SuccBool
% LocalWords:  notstuck unfoldE

>>>>

We have presented in a previous paper~\cite{blanco19ppdp} a
proof-theoretical reconstruction of property-based testing (PBT)~\cite{fink97sen}
of
relational specifications, adopting techniques from foundational proof
certificates to account for several features of this testing
paradigm: from various \emph{generation} strategies, to
\emph{shrinking} and fault localization.

@@

While nothing prevents us from porting all the PBT features we have
accounted for in~\cite{blanco19ppdp}, for the sake of this paper we
will implement only FPC corresponding to different flavors of
\emph{exhaustive} generation, as adopted, e.g., in
SmallCheck~\cite{smallcheck} and
$\alpha$Check~\cite{cheney_momigliano_2017}, and their
combination. Note however that it would take  no more than two lines of
code in the \lsti{decideE} expert to implement a form of random data
generation in the sense of randomized
backtracking~\cite{pltredexconstraintlogic}.

@@

% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat DM
% LocalWords:  tif ttrue tzero tsucc SUT PBT SmallCheck QuickChick
% LocalWords:  decidability shrinkers dep pbt fpc interp forall ness
% LocalWords:  ord unfoldings pre relationally shrinker booleans typ
% LocalWords:  tfalse tpred tiszero TBool TNat subexp SuccBool
% LocalWords:  notstuck unfoldE decideE

### 2020-11-07 papers/types2020/sec3.tex
happy about sec 3 and 4

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
  \item The inference rules are also annotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across the Elpi API with Coq.
\end{enumerate}
The proof system described in Figure~\ref{fig:augmented} (and
implemented in Figure~\ref{fig:kernel}) corresponds to the Calculus
of Inductive Constructions in which all inductive definitions are
limited to the Horn clause.
%
This proof system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
% %
% In our case, the \lP goal formula \lsti{(check Xi (go B T))}
% corresponds to $\XXi\vdash B \colon T$ and the formula
% %
% \lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
% D\vdash B$.
%
This style of  proof terms coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$, which
here we take as a set of inductive definitions,
which, following Coq's reference manual, we denote with
$\mathrm{Ind}[p]  (\Gamma_I := \Gamma_C)$.  One  rule delegates
to Coq's type checking the enforcement of the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with types
that correspond to Horn clauses, and atomic types are inductively
defined. In fact, we do not have a $\forall$ rule on the right,
although the proof theory would gladly allow it. This means at the
term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.
The \emph{decide} rule, %  is the equivalent in our 
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
the rule selects a clause on which to back-chain on: we lookup the
constructors from an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                               % reification of Coq terms
kind gref   type.                               % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: namely $(x :
B) D$ is encoded by \lsti{prod ``x'' B (x\ D x)}; when, in the
implementation of the product-left rule, we apply $D$ to the variable
\lsti{Tm}, we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to introduce for the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase on it.

@@

As it has been described in \cite{blanco17cade}, it is also possible
to pair together two different proof certificates, defined by two
different FPC definitions, and do the proof checking in parallel.
This means that we can build an FPC that \emph{restricts} proofs
satisfying two FPCs simultaneously.  In particular, the infix
constructor \lsti{<c>} in Figure~\ref{fig:resources} forms the pair of
two proof certificates and the pairing experts for the certificate
\lsti{Cert1 <c> Cert2} simply request that the corresponding experts
also succeed for both \lsti{Cert1} and \lsti{Cert2} Thus, the query
\lsti{check ((qheight 4) <c> (qsize 10)) A} will succeed if there is a
proof of \lsti{A} that has a height less than or equal to 4 while also
being of size less than or equal to 10.

@@

The glue code between Coq-Elpi and the implementation of our calculus
is straightforward: the goal is given to us as a quadruple of a (here
inactive) context, an \emph{evar}, a type (goal) and a list of extra
information. In addition, we supply the certificate  when calling the tactic (it simply is an integer and a string to
identify the ``resource'' FPC we will use in this case). We just need
to call \lsti{check} on the goal, together with the
certificate, in order to obtain a reconstructed proof term. We do not
call the reconstruction directly on the evar because Coq-Elpi
ensures that evars manipulated by \lP are well-typed at all times; since
we cannot guarantee that, as we work with partially reconstructed
term, we  get  around this by an explicit unification.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
% LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
% LocalWords:  FPCs Prolog evars untyped

>>>>

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following fours ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
  \item The inference rules are also annotated by terms structures that
    can be give directly to the Coq kernel for checking.
  \item We have added various premises which are responsible to
    pulling information across  Coq's global environment.
\end{enumerate}
The proof system described in Figure~\ref{fig:augmented} (and
implemented in Figure~\ref{fig:kernel}) corresponds to a subset of the Calculus
of Inductive Constructions in which all inductive definitions are
limited to pure Horn clauses.
%
This  system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
% %
% In our case, the \lP goal formula \lsti{(check Xi (go B T))}
% corresponds to $\XXi\vdash B \colon T$ and the formula
% %
% \lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
% D\vdash B$.
%
This style of  proof terms coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$,
which here we take as a set of inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the inductive type and $\Gamma_C$ the names
and types of its constructors. A rule delegates to Coq's
type checking the enforcement of the well-sortedness of such types. We
do not have a \emph{local context} since we are only dealing with
types that correspond to Horn clauses, and atomic types are
inductively defined. In fact, we do not have a $\forall$ rule on the
right, although the proof theory would gladly allow it. This means at
the term level that we do not have variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors from an \lsti{Inductive} definition from the global
environment, one that matches the atom we wish to backchain on, and then
call the latter judgment. The rules for backchaining are standard,
corresponding to the \emph{init} and the (conflation of the) left introduction rules for
$\forall$ and $\supset$.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of the native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                               % reification of Coq terms
kind gref   type.                               % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply $D$ to the variable
\lsti{Tm}, we get a new formula that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in LJT and PTSC's tradition, the more since proof search in
our application is cut-free.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase.

@@

As it has been described in \cite{blanco17cade}, it is also possible
to pair together two different proof certificates, defined by two
different FPC definitions, and do the proof checking in parallel.
This means that we can build an FPC that \emph{restricts} proofs
satisfying two FPCs simultaneously.  In particular, the infix
constructor \lsti{<c>} in Figure~\ref{fig:resources} forms the pair of
two proof certificates and the pairing experts for the certificate
\lsti{Cert1 <c> Cert2} simply request that the corresponding experts
also succeed for both \lsti{Cert1} and \lsti{Cert2} Thus, the query
\verb+check ((qheight 4) <c> (qsize 10)) A+ will succeed if there is a
proof of \lsti{A} that has a height less than or equal to 4 while also
being of size less than or equal to 10.

@@

The glue code between Coq-Elpi and the implementation of our calculus
is straightforward: the goal is given to us as a quadruple of a (here
inactive) context, an \emph{evar}, a type (goal) and a list of extra
information. In addition, we supply the certificate  when calling the tactic (it is simply  an integer and a string to
identify the ``resource'' FPC we will use in this case). We just need
to call \lsti{check} on the goal, together with the
certificate, in order to obtain a reconstructed proof term. We do not
call the reconstruction directly on the evar because Coq-Elpi
ensures that evars manipulated by \lP are well-typed at all times; since
we cannot guarantee that, as we work with partially reconstructed
term, we  get  around this by an explicit unification.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog CIC init lp
% LocalWords:  Andreoli's basicstyle lprolog sigs Coq's sortedness Ev
% LocalWords:  backchain gref indt indc coq dest unfoldings str fpc
% LocalWords:  evar nat forall ys elpi Qed lc args ljf HOAS unfoldE
% LocalWords:  FPCs Prolog evars untyped decideE
